---
title: 'STAT 365 HW #6'
author: "Arthur Erlendsson"
date: "March 9, 2017"
output: word_document
---
```{r}
library(splines)
bone <- read.csv("bone.csv")
head(bone)
```

```{r}
plot(x = bone$age,y = bone$spnbmd,xlab = "age",ylab = "spnbmd",col = bone$gender)
legend(legend = levels(bone$gender),col = c("black","red"),x = "topright",pch=3)
```

It almost looks as though $spnbmd$ might correlate with puberty - the variance of female $spnbmd$ decreases sooner than male $spnbmd$, which leads to female $spnbmd$ values sitting close to 0 at a lower age than those of males.

Piecewise quadratic:

```{r}
bm <- within(bone[bone$gender == "male",], rm(gender))
bf <- within(bone[bone$gender == "female",], rm(gender))

bmq <- quantile(bm$age,c(.33,.67))
bfq <- quantile(bf$age,c(.33,.67))

# hacky but the minimum wasn't being included in the breaks
bm$agegrp <- cut(bm$age, breaks=c(min(bm$age) - .01,bmq,max(bm$age)))
bf$agegrp <- cut(bf$age, breaks=c(min(bf$age) - .01,bfq,max(bf$age)))

labm <- levels(bm$agegrp)
labf <- levels(bf$agegrp)

bdm <- cbind(lower = as.numeric( sub("\\((.+),.*", "\\1", labm) ),
      upper = as.numeric( sub("[^,]*,([^]]*)\\]", "\\1", labm) ))
bdf <- cbind(lower = as.numeric( sub("\\((.+),.*", "\\1", labf) ),
      upper = as.numeric( sub("[^,]*,([^]]*)\\]", "\\1", labf) ))

bdm <- c(bdm[,1], bdm[nrow(bdm),2])
bdf <- c(bdf[,1], bdf[nrow(bdf),2])

mm_pq <- lm(spnbmd ~ agegrp*poly(age, 2, raw=T), data=bm)
mf_pq <- lm(spnbmd ~ agegrp*poly(age, 2, raw=T), data=bf)

length(coef(mm_pq))
length(coef(mf_pq))

plot(spnbmd ~ age, data = bm, pch=16, main="male")
abline(v=bmq,lwd=2,lty=2,col="blue")
curve(predict(mm_pq, data.frame(agegrp=cut(x, breaks=bdm),age=x)), lwd=2, col=2, add=TRUE)

plot(spnbmd ~ age, data = bf, pch=16, main="female")
abline(v=bfq,lwd=2,lty=2,col="blue")
curve(predict(mf_pq, data.frame(agegrp=cut(x, breaks=bdf),age=x)), lwd=2, col=2, add=TRUE)
```


Continuous piecewise quadratic:
```{r}
mm_cpq <- lm(spnbmd ~ poly(age,2, raw=T) + 
            poly(pmax(I(age-bdm[2]),0), 2, raw=T) +
            poly(pmax(I(age-bdm[3]),0), 2, raw=T), data=bm)
mf_cpq <- lm(spnbmd ~ poly(age,2, raw=T) + 
            poly(pmax(I(age-bdf[2]),0), 2, raw=T) +
            poly(pmax(I(age-bdf[3]),0), 2, raw=T), data=bf)

length(coef(mm_cpq))
length(coef(mf_cpq))

plot(spnbmd ~ age, data = bm, pch=16,main="male")
abline(v=bmq,lwd=2,lty=2,col="blue")
curve(predict(mm_cpq, data.frame(agegrp=cut(x, breaks=bdm),age=x)), lwd=2, col=2, add=TRUE)

plot(spnbmd ~ age, data = bf, pch=16,main="female")
abline(v=bfq,lwd=2,lty=2,col="blue")
curve(predict(mf_cpq, data.frame(agegrp=cut(x, breaks=bdf),age=x)), lwd=2, col=2, add=TRUE)
```


Continous piecewise quadratic w/ continuous first derivative:
```{r}
mm_fd <- lm(spnbmd ~ poly(age,2, raw=T) + pmax(I(age-bdm[2])^2,0) + 
            pmax(I(age-bdm[3])^2,0), data=bm)
mf_fd <- lm(spnbmd ~ poly(age,2, raw=T) + pmax(I(age-bdf[2])^2,0) + 
            pmax(I(age-bdf[3])^2,0), data=bf)

length(coef(mm_fd))
length(coef(mf_fd))

plot(spnbmd ~ age, data = bm, pch=16,main="male")
abline(v=bmq,lwd=2,lty=2,col="blue")
curve(predict(mm_fd, data.frame(agegrp=cut(x, breaks=bdm),age=x)), lwd=2, col=2, add=TRUE)

plot(spnbmd ~ age, data = bf, pch=16,main="female")
abline(v=bfq,lwd=2,lty=2,col="blue")
curve(predict(mf_fd, data.frame(agegrp=cut(x, breaks=bdf),age=x)), lwd=2, col=2, add=TRUE)
```


Cubic spline:
```{r}
mm_cs <- lm(spnbmd ~ poly(age,3, raw=T) + pmax(I(age-bdm[2])^3,0) + 
            pmax(I(age-bdm[3])^3,0), data=bm)
mf_cs <- lm(spnbmd ~ poly(age,3, raw=T) + pmax(I(age-bdf[2])^3,0) + 
            pmax(I(age-bdf[3])^3,0), data=bf)

length(coef(mm_cs))
length(coef(mf_cs))

plot(spnbmd ~ age, data = bm, pch=16,main="male")
abline(v=bmq,lwd=2,lty=2,col="blue")
curve(predict(mm_cs, data.frame(agegrp=cut(x, breaks=bdm),age=x)), lwd=2, col=2, add=TRUE)

plot(spnbmd ~ age, data = bf, pch=16,main="female")
abline(v=bfq,lwd=2,lty=2,col="blue")
curve(predict(mf_cs, data.frame(agegrp=cut(x, breaks=bdf),age=x)), lwd=2, col=2, add=TRUE)
```


I like how the cubic splines look - they're more capable of dealing with the flexibility in the data.  What is a bit concerning is the end effects on the lines - usually one would fix this by using natural splines instead of cubic splines but it's not an option for the problem set.

Let's do cross-validation with cubic splines using five folds:
```{r}
set.seed(665)
bm <- bm[sample(1:nrow(bm), replace = FALSE),]
bf <- bf[sample(1:nrow(bf), replace = FALSE),]
foldm <- as.numeric(cut(1:nrow(bm), 5))
foldf <- as.numeric(cut(1:nrow(bf), 5))

errm <- rep(0,6)
errf <- rep(0,6)

for (i in 2:7) {
  quants <- levels(cut(0:100,i+1))
  quants <- cbind(lower = as.numeric( sub("\\((.+),.*", "\\1", quants) ),
      upper = as.numeric( sub("[^,]*,([^]]*)\\]", "\\1", quants) ))
  quants <- quants[-nrow(quants),2]
  
  bmq <- quantile(bm$age,quants/100)
  bmf <- quantile(bf$age,quants/100)
  
  bm$agegrp <- cut(bm$age, breaks=c(min(bm$age) - .01,bmq,max(bm$age)))
  bf$agegrp <- cut(bf$age, breaks=c(min(bf$age) - .01,bfq,max(bf$age)))
  
  labm <- levels(bm$agegrp)
  labf <- levels(bf$agegrp)
  
  bdm <- cbind(lower = as.numeric( sub("\\((.+),.*", "\\1", labm) ),
      upper = as.numeric( sub("[^,]*,([^]]*)\\]", "\\1", labm) ))
  bdf <- cbind(lower = as.numeric( sub("\\((.+),.*", "\\1", labf) ),
      upper = as.numeric( sub("[^,]*,([^]]*)\\]", "\\1", labf) ))
  
  bdm <- c(bdm[,1], bdm[nrow(bdm),2])
  bdf <- c(bdf[,1], bdf[nrow(bdf),2])
  
  for (f in 1:5) {
    bm_train <- bm[foldm != f,]
    bm_valid <- bm[foldm == f,]
    bf_train <- bf[foldf != f,]
    bf_valid <- bf[foldf == f,]
    
    mm <- lm(spnbmd ~ bs(age, knots=bdm[2:i+1]), data=bm_train)
    mf <- lm(spnbmd ~ bs(age, knots=bdf[2:i+1]), data=bf_train)
    
    pm <- predict(mm, newdata=bm_valid)
    pf <- predict(mf, newdata=bf_valid)
    
    errm[i-1] <- errm[i-1] + sum((bm_valid$spnbmd - pm)^2,na.rm=TRUE)/nrow(bm_valid)
    errf[i-1] <- errf[i-1] + sum((bf_valid$spnbmd - pf)^2,na.rm=TRUE)/nrow(bf_valid)
  }
}

err <- errm + errf
err
num_knots <- which.min(err) + 1
num_knots
```


So let's fit on five knots.
```{r}
bq <- quantile(bone$age,c(.167,.333,.50,.667,.833))

# hacky but the minimum wasn't being included in the breaks
bone$agegrp <- cut(bone$age, breaks=c(min(bone$age) - .01,bq,max(bone$age)))

labs <- levels(bone$agegrp)

bds <- cbind(lower = as.numeric( sub("\\((.+),.*", "\\1", labs) ),
      upper = as.numeric( sub("[^,]*,([^]]*)\\]", "\\1", labs) ))

bds <- c(bds[,1], bds[nrow(bds),2])

m_opt <- lm(spnbmd ~ bs(age, knots=bds[2:6]), data=bone)

plot(spnbmd ~ age, data = bone, pch=16,main="all data")
abline(v=bq,lwd=2,lty=2,col="blue")
curve(predict(m_opt, data.frame(agegrp=cut(x, breaks=bds),age=x)), lwd=2, col=2, add=TRUE)
```